Making multiple-locus association mapping on a genome-wide scale routine

### 2016-09-06 AM+draft.tex
06/09 Playing around with the intro (first paragraph). Have near finalised abstract.

\abstract{
In this paper, we present our method that makes multiple locus association mapping routine. 
It is only a little more complicated than performing a linear mixed model analysis. 
Yet, we are able to fit all marker-trait associations simultaneously while still clearly identifying
 which marker-trait associations are true. There are no thresholds to be set, no multiple 
 testing issues,  and no parameters to fine tune. We will show that our method is capable 
 of  finding marker-trait associations more accurately,  and in many cases, orders of 
 magnitude faster than competing approaches. Our multiple locus method is implemented 
 in a freely available, fully documented and maintained R package called am+.
}

@@

Gwas for many are now the design of choice when interested in elucidating the genetic 
underpinnings  of complex traits.  Over the past decade, gwas have changed 
considerably in focus and scope. Early gwas 
were performed predominantly as case control studies. Their focus 
was on identifying the casual variants underlying common disorders in humans. 
Despite many promising findings, reproducibility of these findings 
proved an issue (refs). With the collection of dense whole genome data challenging, 
early gwas lacked power (refs).  Today, gwas are performed not just in humans but 
animals and plants that may have highly structured pedigrees. 
They can involve hundreds to thousands of individuals with genotypes collected 
from thousands to hundreds of thousands of snps. Where early gwas were interested
 in disease traits, modern gwas are also  interested in  categorical and quantitative traits.

Over the past decade, there have also been impressive advancements in genome wide 
association mapping methods, the statistical techniques for measuring the strength of 
association between a marker locus and trait. Traditionally, association data were 
analysed with simple contingency tests or Cochran-armitage  trend tests. Today, far 
more statistically sophisticated methods are available. One such method is the  QK 
approach, a linear mixed model framework for genome wide association 
mapping (refs).  Several highly efficient computer programs for performing QK-based 
association mapping are available. Programs such as EMMA, fastLMM, GEMMA, and 
XXX(ref) make feasible the analysis of data from hundreds of thousands of snps, collected 
on thousands of subjects. However, not all areas of genome-wide association 
mapping have advanced. It still remains standard practice to analyse genome-wide 
association data with a single locus model. That is,  it is common for the strength of 
association between a marker locus and trait is measured for each locus separately.

>>>>

\abstract{
The methods of choice for genome wide association mapping all share a common shortcoming. 
They measure the strength of marker-trait associations on a locus by locus basis. 
More sophisticated multi-locus methods that map multiple marker-trait associations jointly are available. 
However, these methods are computationally demanding, are not always easy to implement, and can give results that are difficult to interpret. The aim of this paper is to present our method that makes the multi-locus analysis of data from genome wide association studies routine. It is formulated within a linear mixed model framework. We call our method AMplus. AMplus produces results faster than competing multi locus methods and often with greater statistical power. In terms of speed, AMplus is just a little slower than the fastest single-locus linear mixed model implementations. AMplus is freely available as a fully documented R package which we port as a docker binary for platform independent portability.
}

@@

Over the past decade, gwas have changed 
considerably in scope and method of analysis. Early gwas were performed as case-control studies and their data analysed with simple
contingency table tests or logistic regressions.  Differences in marker allele frequency between cases (affecteds) and controls (unaffecteds) was assumed to be due to a disease variant. It was soon realised though that 
unrecognised population stratification had the capacity to cause spurious associations with non-causal genes. This led to a rise in 
popularity for family-based designs. Family-based designs are robust to population stratrification. 
Consequently, there has been a myriad of association studies performed with affected sibling pairs, 
parent-offspring trios, and nuclear families. All have an accompanying plethora of parametric (refs) and non-parametric (refs) analysis methods. However, as the popularity of GWAS as the study design of choice grew, so did the desire to perform GWAS in general 
populations which may have complex, possibly unknown, structure. Here, we could no longer rely on study design to control population stratification. There began a push to develop methods which were capable of accounting for the confounding affects of population 
stratification. Early methods include genomic control and structured association. However, it has been development of a linear mixed 
modelling framework for GWAS that has the greatest flexibility for controlling for popuation structure and familial relatedness in general population studies.

so to did the desire to implement GWAS in species with structured populations butw where the family structure is unknown

 the statistical significance of differences in marker allele frequency determined with 
contingency table tests or logistic regression. It was soon realised though that unrecognised population heterogeniety 
had the potential to confound findings. This led to the rise in popularity of family-based designs based on 
sibling pairs, trios, and nuclear families.  These designs are robust to deleterious affects of population stratification.

 along with a plethora of
different score and frequency tests being developed.

and 
and their data analysed either 
with simple contingency table tests or logistic regression. The deleterious affects of hidden population heterogeniety though on findings was soon recognised. 
This led to the preponderance of family-based studies because they are robust to potential population heterogeniety.

 Whether performed with trios, nuclear families, 
sibling pairs, or extended families, family-based studies correctly controlled population stratification.

controlling for spurious associations 
due to latent population heterogeniety soon became a focus. This led to family based designs becoming popular. These designs

 based on 
trios,

involving trios, nuclear families, sibling pairs, or ext

were performed predominantly as case control or family based studies. Their focus 
was on identifying the casual variants underlying common disorders in humans. As the popularity of GWAS grew, 
they started to be performed in populations of unrelated individuals and even in structured populations with complicated pedigrees. 
Here, the traits of interest also included quantitative traits. Today, gwas are being performed at the population level, with studies 
involving thousands of individuals with genotypes collected from hundreds of thousands of snps.

Gwas for many are now the design of choice when interested in elucidating the genetic 
underpinnings  of complex traits.  Over the past decade, gwas have changed 
considerably in focus and scope. Early gwas 
were performed predominantly as case control or family based studies. Their focus 
was on identifying the casual variants underlying common disorders in humans. 
Despite many promising findings, reproducibility of these findings 
proved an issue (refs). Early gwas  lacked statistical power because of the 
challenges of collecting dense whole genome data across a sufficiently large number of samples (refs).
Today, gwas are being performed in a wide variety of species. These studies can involve thousands
 of individuals with genotypes collected from thousands to hundreds of thousands of snps. 
Where early gwas were interested
 in disease traits, modern gwas are also  interested in  categorical and quantitative traits.

Over the past decade, there have also been impressive advancements in genome wide 
association mapping methods, the statistical techniques for measuring the strength of 
association between a marker locus and trait. (Needs more detail - what are some of the early assoc study methods (give a single sentence review).... TDT, genomic control, structured association, score tests???? 
Traditionally, association data were 
analysed with simple contingency tests or Cochran-armitage  trend tests. Today, far 
more statistically sophisticated methods are available. One such method is the  QK 
approach, a linear mixed model framework for genome wide association 
mapping (refs).  The QK approach

Several highly efficient computer programs for performing QK-based 
association mapping are available. Programs such as EMMA, fastLMM, GEMMA, and 
XXX(ref) make feasible the analysis of data from large GWAS. 
 However, not all areas of genome-wide association 
mapping have advanced. It still remains standard practice to analyse genome-wide 
association data with a single locus model. That is,  it is common for the strength of 
association between a marker locus and trait is measured for each locus separately.
